Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-08-23 paper.Rnw
Fully responded to Matt's comments

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
\newcommand{\FEPdataset}{SZ-FEP }
\newcommand{\gs}{Gold Standard }

@@

\paragraph{Evaluation method}

The Dice similarity coefficient (DSC), also known as Dice's Kappa, assesses the
agreement between two segmentations. It is one of the most widely used measures
of segmentation accuracy, and we use it as the basis of comparison in this
experiment.
%Additionally, we report the Jaccard index, another commonly used similarity
%measure:

@@

\paragraph{Evaluation method}
The manual and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated \citep{Winterburn2013}. This
difference poses a problem for evaluation by measuring overlap.  That is, since
different protocols will necessarily produce segmentations that do not perfectly
overlap, the degree of overlap cannot be solely used to compare segmentation
methods using different protocols. In place of an overlap metric, we can assess
the degree of correlation in average hippocampal volume of the subjects produced
by each method. Specifically, a Pearson correlation is used.

@@

\paragraph{Evaluation method}
As in Experiment 3, the SNT \gs and Winterburn hippocampal segmentation
protocols differ in the neuroanatomical features delineated, and so we must
assess \mb by the degree of correlation of average hippocampal volume across all
subjects. Specifically, a Pearson correlation is used. We also compute the
correlation in hippocampal volume between the existing, established automated
segmentation methods: FSL, FreeSurfer, and MAPER.  Additionally, we evaluate the
volume-related fixed and proportional biases in all segmentation methods using
Bland-Altman plots \citep{Bland1986}.

@@

In this experiment we conducted 10 rounds of \mb and multi-atlas segmentation of
each of 69 subjects at a range of atlas and template library sizes, registration
algorithm (\ants or \animal), and three label fusion techniques.  Hippocampal
\mb-based segmentations using both \animal and \ants registration algorithm
demonstrate good overlap with SNT derived gold-standards (Figure
\ref{fig:ADNI1-xval-k-mean}). Qualitatively, both \animal and \ants-based
segmentations demonstrate trend overlap accuracy that increases with the size of
atlas library and template library. Improvement in accuracy diminishes noticeably
with template libraries larger than roughly ten images.

No marked difference in segmentation accuracy is seen when either \animal or \ants
registration is used with any number of atlases or templates.  In every parameter 
configuration, the use of \mb with \ants registration shows a pronounced increase
in segmentation accuracy over \mb with \animal registration.  Surprisingly, the 
label fusion method used does not significantly improve label accuracy, contrary 
to the findings of \citet{Aljabar2009} when using weighted voting on much larger
atlas/template libraries. In the remainder of  this section, only results using
the \ants registration algorithm and majority vote fusion will be shown.

With an increasing number of templates, \mb shows improvement in overlap
accuracy over multi-atlas-based segmentation when using the same number of atlases
and voting method (Figure \ref{fig:ADNI1-xval-k-diff}). The magnitude of improvement over
multi-atlas-based segmentation decreases with an increasing number of atlases,
with accuracy converging with 7 atlases.  Peak improvement in \mb accuracy (~0.02
DSC) is found when one atlas is used with a template library of 20 images.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-xval-prep,dependson='ADNI1-xval-load',include=FALSE,cache=TRUE>>=
all_data  <- read.csv(gzfile('data/cache/ADNI-XVAL:all_data.csv.gz'))
all_data_mean  <- read.csv(gzfile('data/cache/ADNI-XVAL:all_data_mean.csv.gz'))
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-prep', fig.width=7, fig.height=7>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(all_data_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=as.factor(atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb segmentations with 
  SNT \gs segmentations for 69 ADNI1 subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. Points above zero
  indicate an \mb parameter settings yielding a higher mean DSC score than
  multi-atlas segmentation using the same number of atlases.
  \label{fig:ADNI1-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: "increase" in mean kappa over multi-atlas
\begin{figure}
<<fig:ADN11-xval-k-diff,cache=TRUE,dependson='ADNI1-xval-prep'>>=
ggplot(subset(all_data_mean, reg_method=="ANTS" & 
                method.mb=="Majority Vote" & (templates.mb*atlases) %% 2 == 1), 
       aes(x=templates.mb, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_summary(fun.y=mean, geom="line", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,2)) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Number of Atlases") +
  scale_alpha_continuous(guide = "none") + 
  scale_linetype_discrete(guide = "none") + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference in mean DSC" ) +
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Difference in mean Dice's Similarity Coefficient of \mb and
  multi-atlas segmentations with \gs (SNT) segmentations for a range of atlas
  and template library sizes. \label{fig:ADNI1-xval-k-diff}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference of variance during validation (p-value)") +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variance of segmentation accuracy between \mb and multi-atlas segmentation 
  is computed for each subject across all ten rounds of validation. Shown on the 
  y-axis (scaled logarithmically) is the  p-value resulting from a t-test 
  comparing the distribution of variances at each parameter setting 
  (atlas/template library size). Only points where \mb mean variability is lower 
  than multi-atlas are shown.}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

ggplot(melted, aes(y=percenterr,x=region,colour=measure)) +
  geom_hline(aes(yintercept=0), colour="grey", size=1) + 
  geom_boxplot() +
  labs(y=expression(paste("Percent error in volume relative to 0.3", mm^3, " Winterburn Atlases")), x="Region") +
  scale_colour_discrete("") + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Percent error in segmentation volume by hippocampus subregion.} 
  Percent error is measured against the volumes of of the unmodified Winterburn
  atlas segmentations.
  {\bf Subsampled} are volumes of the manual segmentations of the Winterburn
  atlases after resampling to $0.9mm^3$.  
  {\bf MAGeT + WA Subsampled} volumes are \mb segmentations of the Winterburn atlas
  images after resampling to $0.9mm^3$ voxels.
  {\bf MAGeT + WA BRAVO} volumes are \mb segmentations of T1 BRAVO images
  ($0.9mm^3$ voxels) acquired separately of four of the five Winterburn atlas
  subjects.} 
  \label{fig:WAval-vol-boxplot}
\end{figure}

@@

To investigate the level of agreement with SNT \gs hippocampal
volumes, we constructed Bland-Altman plots for each method (Figure
\ref{fig:ADNI-Bland-Altman}).  As \citet{Bland1986} noted, high correlation
amongst measures of the same quantity does not necessarily imply agreement (as
correlation can be driven by a large range in true values, for instance).  What
is most striking in Figure \ref{fig:ADNI-Bland-Altman} is that all methods show
an obvious proportional bias: FreeSurfer and FSL markedly under-estimate smaller
hippocampi and over-estimate large hippocampi, whereas MAPER and \mb more
conservatively show the reverse bias.  Additionally, all methods show a fixed
bias, with FreeSurfer and FSL most dramatically over-estimating hippocampal
volume by $2600 mm^3$ and $2800 mm^3$ on average, respectively, and MAPER and
\mb within $250 mm^3$ on average.

@@

qplot(DX,value,data=m2,
      colour=Method,geom="boxplot") + 
      #geom_text(aes(y=-Inf,x=c(.7,.85,1,1.15, 1.7,1.85,2,2.15, 2.7,2.85,3,3.15), vjust=-5, label=round(pearson,2)), 
      #          colour = 'black', size=3, data=correlations) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Hippocampal volume (", mm^3, ")"))) + 
      theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset
dataset by FreeSurfer, FSL, MAPER, \mb, and SNT \gs vs. disease
category.}
  \label{fig:ADNI-volumes-boxplot}
\end{figure}

@@

ggplot(data=melted, aes(x=SNT, y=volume, colour=Method)) + 
    geom_point(size=1) + 
    geom_smooth(method="lm") + 
    xlab(expression("SNT \\gs hippocampal volume " (mm^3))) + 
    ylab(expression("Automated hippocampal volume " (mm^3))) +
    annotate("text", y = c(2000,1500,1000,500), x=Inf-100, size=2.5,
              label = c(paste("FS r =", round(cor(df$SNT, df$FS),2)),
                        paste("FSL r =", round(cor(df$SNT, df$FSL),2)),
                        paste("MAPER r =", round(cor(df$SNT, df$MAPER),2)),
                        paste("MAGeT r =", round(cor(df$SNT, df$MAGeT),2))),
              hjust=1.1,vjust=1.5) +
    theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset
dataset by each of the four automated methods investigated (FreeSurfer (FS),
FSL, MAPER, \mb) vs. SNT \gs. Linear fit lines and pearson
correlations are shown for each method.}
  \label{fig:ADNI-volumes-plot}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Tabulation of segmentation accuracies of existing methods (ADNI)
%
%
<<mb-best-kappa, echo=F, cache=FALSE>>=
ants.mean = subset(all_data_mean, reg_method=="ANTS" & method.mb=="Majority Vote" &      templates.mb == 19 & atlases == 9)
mb.best.kappa = mean(ants.mean$k.mb)
@

\begin{table}[!tbp]
\scriptsize
\caption{\textbf{Automated segmentation accuracy (overlap with SNT \gs labels)
of the ADNI dataset.} For each method, the number of labelled atlases
used for training, the best Dice's overlap measure, the disease classes
measured, and the validation procedure are shown. Unless specified, validation
datasets are composed equally of subjects diagnosed with Alzheimer's Disease
(AD), Mild Cognitive Impairment (MCI), and Cognitively Normal (CN). LOOCV =
Leave-one-out cross-validation. 
   Some studies of automated segmentation of ADNI images are excluded because they 
   do not not provide overlap measures for the hippocampus \cite{Heckemann2011,Chupin2009}.
  \label{tab:other-methods}} 
\begin{center}
\begin{tabular}{ l  c  p{1in}  l  p{2in}  }
\hline\hline
Method & Atlases & DSC & Reference & Validation \tabularnewline
\hline
\mb                     &9 & \Sexpr{round(mb.best.kappa, digits=3)} & & 10 rounds of Monte Carlo CV on a pool of 69 subjects \tabularnewline
LEAP                    &30&0.848                 &\cite{Wolz2010}  &Segmentation of 60 subjects\tabularnewline
ACM (AdaBoost-based)    &21&0.862                 &\cite{Morra2008} &LOOCV on atlases\tabularnewline
Patch-based label fusion&16&0.883 (CN) \newline 0.838 (AD)&\cite{Coupe2011e}&LOOCV on atlases\tabularnewline
Multi-atlas             &30&0.885                 &\cite{Lotjonen2010}&Segmentation of 60 subjects\tabularnewline
Multi-atlas + weighted fusion&20&0.898 (CN) \newline 0.798 (left HC, MCI) &\cite{Wang2011}  &10 rounds of Monte Carlo CV on 20 subjects, pool of 139 (CN/MCI)\tabularnewline
Multi-atlas (MAPS)    &55&0.890                   &\cite{Leung2010}&Segmentation of 30 subjects (10 AD, MCI, and CN)\tabularnewline
\hline
\end{tabular}
\end{center}
\end{table}

@@

Throughout the cross validation in Experiment 1 (10-fold Monte Carlo cross 
validation in the
\adnidataset dataset subsample) we find that two parameter choices improve 
segmentation accuracy: increasing the number of atlases, and the number of
templates. However, after setting the parameters to 5 atlases and 15 templates
there are diminishing returns with respect to this improvement. We were
somewhat surprised to find that cross-correlation and normalized mutual
information based weighted voting did little to improve segmentation accuracy.
This suggests that using an intermediate template library generated by accurate
nonlinear registration methods such as \ants \citep{Avants2008} sufficiently filters
sources of error such as nonlinear registration error and partial volume effects
that arise through the use of nearest neighbour resampling. From our previous
work on the \mb algorithm we have shown that these increases in error
are not simply a smoothing or averaging effect \citep{MallarChakravarty2012}.

@@

To this end, more work is required in order to determine the source of the
slight decrease in segmentation performance when the number of templates are
set to an even number.  Our initial concern was that this dip in performance
was a by-product of the \mb algorithm itself.  However, this this pattern is
also found in the results of the multi-atlas segmentations we used in our
experiments.  We believe that our majority voting methodology is biased towards
labels with the lowest numeric values when breaking ties (by way of the
implementation of the \verb+mode+ function used to determine majority), thus
causing the slight bias observed when using an even number of templates. This
is another area where the voting scheme could be used to improve performance.
However, it is worth noting that this limitation was previously identified by
\citet{Heckemann2006a} and, subsequently, other groups have not even considered
the potential pitfalls of an even number of candidate labels (eg.
\citet{Leung2010}).

@@

In comparison to other methodologies in the field, it is important to note that
\mb performs quite well.  Table 6 gives a survey of some of the most recent
multi-atlas implementations with the reported kappa values used to validate the
algorithm on the ADNI dataset (many of which use the same SNT labels for
populating their template library and as gold standards for evaluation).  While
it is difficult to compare segmentation results across studies, gold-standards,
evaluation metrics, and algorithms it is worth noting that the methods
summarized in Table 6 require more atlases (between 16-55) than our \mb
implementation with the Winterburn atlases \citep{Winterburn2013}.  Amongst these
methods only method 4 yields mean Kappas that are higher than the ones that are
reported here.

>>>>

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
\newcommand{\FEPdataset}{SZ-FEP }
\newcommand{\gs}{Gold Standard }
\newcommand{\fsl}{FSL }
\newcommand{\freesurfer}{FreeSurfer }
\newcommand{\maper}{MAPER }

@@

\paragraph{Evaluation method}  
The Dice similarity coefficient (DSC), also known as Dice's Kappa, assesses the
agreement between two segmentations. It is one of the most widely used measures
of segmentation accuracy, and we use it as the basis of comparison in this
experiment.
%Additionally, we report the Jaccard index, another commonly used similarity
%measure:

@@

\paragraph{Evaluation method}
The Pruessner and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated \citep{Winterburn2013}. This
difference poses a problem for evaluation by measuring overlap.  That is, since
different protocols will necessarily produce segmentations that do not perfectly
overlap, the degree of overlap cannot be solely used to compare segmentation
methods using different protocols. In place of an overlap metric, we can assess
the degree of correlation in average hippocampal volume of the subjects produced
by each method. Specifically, a Pearson correlation is used.

@@

\paragraph{Evaluation method}
As in Experiment 3, the SNT \gs and Winterburn hippocampal segmentation
protocols differ in the neuroanatomical features delineated, and so we 
assessed \mb by the degree of (Pearson) correlation of average hippocampal volume
across subjects. We also computed the correlation in hippocampal volume between
the existing, established automated segmentation methods: FSL, FreeSurfer, and
MAPER. Additionally, we evaluate the volume-related fixed and proportional
biases in all segmentation methods using Bland-Altman plots \citep{Bland1986}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-xval-prep,dependson='ADNI1-xval-load',include=FALSE,cache=TRUE>>=
all_data  <- read.csv(gzfile('data/cache/ADNI-XVAL:all_data.csv.gz'))
all_data_mean  <- read.csv(gzfile('data/cache/ADNI-XVAL:all_data_mean.csv.gz'))

mb.best.kappa = mean(subset(all_data_mean, reg_method   == "ANTS" & 
                                  method.mb    == "Majority Vote" &
                                  templates.mb == 19 & 
                                  atlases      == 9)$k.mb)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this experiment we conducted 10 rounds of \mb and multi-atlas segmentation of
of 69 subjects using atlas and template library sizes varying from 1-9 and 1-20
images respectively, two registration algorithms (\ants or \animal), and three
label fusion techniques (unweighted, cross-correlation, and normalised mutual
information weighted majority vote).  Hippocampal \mb-based segmentations using
both \animal and \ants registration algorithm demonstrate good overlap with SNT
\gs segmentations (maximum mean DSC of \Sexpr{round(mb.best.kappa, digits=3)}
when using 9 atlases, 19 templates, ANTS registration, and majority vote label
fusion); Figure \ref{fig:ADNI1-xval-k-mean}).  Qualitatively, both \animal and
\ants-based segmentations demonstrate trend overlap accuracy that increases with
the size of atlas library and template library. Improvement in accuracy plateaus
with template libraries larger than ten images.

The use of \mb with \ants registration shows a pronounced increase
in segmentation accuracy over \mb with \animal registration, across all
other variable settings we tested. Additionally, by itself, using a weighted
voting strategy did not significantly improve segmentation accuracy, contrary to
the findings of \citet{Aljabar2009} in basic multi-atlas segmentation. Given
these findings, in the remainder of this section only results using the \ants
registration algorithm and majority vote fusion will be shown.

With an increasing number of templates, \mb shows improvement over
multi-atlas-based segmentation in overlap accuracy when using the same number of
atlases and voting method (Figure \ref{fig:ADNI1-xval-k-diff}). The two methods
converge in accuracy when using seven atlases.  Peak improvement in \mb accuracy
(~0.02 DSC) is found when one atlas is used with a template library of 19
images.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-prep', fig.width=7, fig.height=7>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(all_data_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=as.factor(atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  opts(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb segmentations with 
  SNT \gs segmentations for 69 ADNI1 subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. Points above zero
  indicate \mb parameter settings yielding a higher mean DSC score than
  multi-atlas segmentation using the same number of atlases.
  \label{fig:ADNI1-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: "increase" in mean kappa over multi-atlas
\begin{figure}
<<fig:ADN11-xval-k-diff,cache=TRUE,dependson='ADNI1-xval-prep'>>=
ggplot(subset(all_data_mean, reg_method=="ANTS" & 
                method.mb=="Majority Vote" & (templates.mb*atlases) %% 2 == 1), 
       aes(x=templates.mb, y=k_diff, colour=as.factor(atlases))) + 
  #facet_grid(reg_method~method.y) + 
  stat_summary(fun.data=mean_cl_boot,geom='errorbar',
               aes(y=k_diff,colour=as.factor(atlases), width=0.2)) +
  stat_summary(fun.y=mean, geom="point", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  stat_summary(fun.y=mean, geom="line", 
               aes(y=k_diff,colour=as.factor(atlases))) + 
  #stat_smooth(method=lm, formula=y~log(x)) + 
  geom_hline(aes(alpha=0.5, yintercept=0), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,2)) + scale_y_continuous(breaks=seq(-1,1,by=0.01)) + 
  scale_colour_hue(name="Number of Atlases") +
  scale_alpha_continuous(guide = "none") + 
  scale_linetype_discrete(guide = "none") + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference in mean DSC" ) +
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Difference in mean Dice's Similarity Coefficient of \mb and
  multi-atlas segmentations with \gs (SNT) segmentations for a range of atlas
  and template library sizes. \label{fig:ADNI1-xval-k-diff}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=1) +
  geom_point(size=3) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Difference of variance during validation (p-value)") +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variance of segmentation accuracy between \mb and multi-atlas segmentation 
  is computed for each subject across all ten rounds of validation. Shown on the 
  y-axis (scaled logarithmically) is the  p-value resulting from a t-test 
  comparing the distribution of variances at each parameter setting 
  (atlas/template library size). Only points where \mb mean variability is lower 
  than multi-atlas are shown.}
  \label{ADNI1-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

ggplot(melted, aes(y=percenterr,x=region,colour=measure)) +
  geom_hline(aes(yintercept=0), colour="grey", size=1) + 
  geom_boxplot() +
  labs(y="Percent error in volume", x="Subfield") +
  scale_colour_discrete("") + 
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Average percent error in hemispheric subfield volume.} 
  Percent error is measured against the volumes of the unmodified Winterburn
  atlas subfield segmentations.
  {\bf Subsampled} are volumes of the manual segmentations of the Winterburn
  atlases after resampling to $0.9mm^3$.  
  {\bf MAGeT + Subsampled} volumes are \mb segmentations of the Winterburn atlas
  images after resampling to $0.9mm^3$ voxels.
  {\bf MAGeT + BRAVO} volumes are \mb segmentations of T1 BRAVO images
  ($0.9mm^3$ voxels) acquired separately of four of the five Winterburn atlas
  subjects.} 
  \label{fig:WAval-vol-boxplot}
\end{figure}

@@

To investigate the level of agreement with SNT \gs hippocampal
volumes, we constructed Bland-Altman plots for each method (Figure
\ref{fig:ADNI-Bland-Altman}).  As \citet{Bland1986} noted, high correlation
amongst measures of the same quantity does not necessarily imply agreement (as
correlation can be driven by a large range in true values, for instance).  All
methods show an obvious proportional bias: FreeSurfer and FSL markedly
under-estimate smaller hippocampi and over-estimate large hippocampi, whereas
MAPER and \mb show a reverse, conservative bias (Figure
\ref{fig:ADNI-Bland-Altman}).  Additionally, all methods show a fixed volume
bias, with FreeSurfer and FSL most dramatically over-estimating hippocampal
volume by $2600 mm^3$ and $2800 mm^3$ on average, respectively, and MAPER and
\mb within $250 mm^3$ on average.

@@

qplot(DX,value,data=m2,
      colour=Method,geom="boxplot") + 
      #geom_text(aes(y=-Inf,x=c(.7,.85,1,1.15, 1.7,1.85,2,2.15, 2.7,2.85,3,3.15), vjust=-5, label=round(pearson,2)), 
      #          colour = 'black', size=3, data=correlations) +
      xlab("Diagnosis") + 
      ylab(expression(paste("Hippocampal volume (", mm^3, ")"))) + 
      opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset
dataset by FreeSurfer, FSL, MAPER, \mb, and SNT \gs vs. disease
category.}
  \label{fig:ADNI-volumes-boxplot}
\end{figure}

@@

ggplot(data=melted, aes(x=SNT, y=volume, colour=Method)) + 
    geom_point(size=1) + 
    geom_smooth(method="lm") + 
    xlab(expression("SNT \\gs hippocampal volume " (mm^3))) + 
    ylab(expression("Automated hippocampal volume " (mm^3))) +
    annotate("text", y = c(2000,1500,1000,500), x=Inf-100, size=2.5,
              label = c(paste("FS r =", round(cor(df$SNT, df$FS),2)),
                        paste("FSL r =", round(cor(df$SNT, df$FSL),2)),
                        paste("MAPER r =", round(cor(df$SNT, df$MAPER),2)),
                        paste("MAGeT r =", round(cor(df$SNT, df$MAGeT),2))),
              hjust=1.1,vjust=1.5) +
    opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset
dataset by each of the four automated methods investigated (FreeSurfer (FS),
FSL, MAPER, \mb) vs. SNT \gs. Linear fit lines and pearson
correlations are shown for each method.}
  \label{fig:ADNI-volumes-plot}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Tabulation of segmentation accuracies of existing methods (ADNI)
%
%
\begin{table}[!tbp]
\scriptsize
\caption{\textbf{Automated segmentation accuracy (overlap with SNT \gs labels)
of the ADNI dataset.} For each method, the number of labelled atlases
used for training, the best Dice's overlap measure, the disease classes
measured, and the validation procedure are shown. Unless specified, validation
datasets are composed equally of subjects diagnosed with Alzheimer's Disease
(AD), Mild Cognitive Impairment (MCI), and Cognitively Normal (CN). LOOCV =
Leave-one-out cross-validation. 
   Some studies of automated segmentation of ADNI images are excluded because they 
   do not not provide overlap measures for the hippocampus \cite{Heckemann2011,Chupin2009}.
  \label{tab:other-methods}} 
\begin{center}
\begin{tabular}{ l  c  p{1in}  l  p{2in}  }
\hline\hline
Method & Atlases & DSC & Reference & Validation \tabularnewline
\hline
\mb                     &9 & \Sexpr{round(mb.best.kappa, digits=3)} & & 10 rounds of Monte Carlo CV on a pool of 69 subjects \tabularnewline
LEAP                    &30&0.848                 &\cite{Wolz2010}  &Segmentation of 60 subjects\tabularnewline
ACM (AdaBoost-based)    &21&0.862                 &\cite{Morra2008} &LOOCV on atlases\tabularnewline
Patch-based label fusion&16&0.883 (CN) \newline 0.838 (AD)&\cite{Coupe2011e}&LOOCV on atlases\tabularnewline
Multi-atlas             &30&0.885                 &\cite{Lotjonen2010}&Segmentation of 60 subjects\tabularnewline
Multi-atlas + weighted fusion&20&0.898 (CN) \newline 0.798 (left HC, MCI) &\cite{Wang2011}  &10 rounds of Monte Carlo CV on 20 subjects, pool of 139 (CN/MCI)\tabularnewline
Multi-atlas (MAPS)    &55&0.890                   &\cite{Leung2010}&Segmentation of 30 subjects (10 AD, MCI, and CN)\tabularnewline
\hline
\end{tabular}
\end{center}
\end{table}

@@

Throughout the cross validation in Experiment 1 (10-fold Monte Carlo cross
validation in the \adnidataset dataset subsample) we find that two parameter
choices improve segmentation accuracy: increasing the number of atlases, and the
number of templates. However, after setting the parameters to 5 atlases and 15
templates there are diminishing returns with respect to this improvement.

Previous work on multi-atlas segmentation methods
\citep{Aljabar2009,Collins2010} has found that cross-correlation and normalized
mutual information based weighted label fusion improves segmentation accuracy
over simple majority vote label fusion. Selectively filtering out atlases with
lower image similarity is believed to reduce sources of error from estimating
deformations via nonlinear registration, partial volume effects from nearest
neighbour image resampling, and neuroanatomical mismatch between atlases and
subjects. That \mb does not see the same boost in performance from weighted
voting may suggest that the neuroanatomical variability of a template library
constructed from study subjects more closely matches any particular subject and
thereby leaving less error to filter. From our previous work on the \mb
algorithm we have shown that the reduction in error is not simply a smoothing or
averaging effect \citep{MallarChakravarty2012}.

@@

To this end, more work is required to determine the source of the slight
decrease in segmentation performance when the number of templates are set to an
even number.  Our initial concern was that this dip in performance was a
by-product of the \mb algorithm itself.  However, this this pattern is also
found in the results of the multi-atlas segmentations we used in our
experiments.  We believe that our majority voting methodology is biased towards
labels with the lowest numeric values when breaking ties (by way of the
implementation of the \verb+mode+ function used to determine majority), thus
causing the slight bias observed when using an even number of templates. This is
another area where the voting scheme could be used to improve performance.
However, it is worth noting that this limitation was previously identified by
\citet{Heckemann2006a} and, subsequently, other groups have not even considered
the potential pitfalls of an even number of candidate labels (eg.
\citet{Leung2010}).

@@

In comparison to other methodologies in the field \mb performs favourably. Table
\ref{tab:other-methods} surveys some of the most recent reported kappa values on
the ADNI dataset using the same SNT labels for populating their template library
and as gold standards for evaluation.  While it is difficult to compare
segmentation results across studies, gold-standards, evaluation metrics, and
algorithms it is worth noting that the methods summarized in Table
\ref{tab:other-methods} require more atlases (between 16-55) than our \mb
implementation with the Winterburn atlases \citep{Winterburn2013}.
